---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Genome-Wide Association Study (GWAS)

```{r setup, include=FALSE}
WORDS_TO_IGNORE <- c("gwas", "HDL", "https", "LDL", "lipoprotein", "batchtools", "Slurm", "ACATO", "ACATV", "Conda", "GxE", "GxG", "Regeneron", "regenie", "REGENIE", "REMETA", "SBAT", "SKAT", "SKATO", "COJO", "GCTA", "GC", "HPC", "iteratively")
source("knitr-options.R")
source("spelling-check.R")
```

In bigstatsr, you can perform both standard linear and logistic regressions GWAS, using either `big_univLinReg()` or `big_univLogReg()`.
Function `big_univLinReg()` should be very fast, while `big_univLogReg()` is slower.

:::: {.infobox .info}
This type of association, where each variable is considered independently, can be performed for any type of FBM (i.e. it does not have to be a genotype matrix). This is why these two functions are in package bigstatsr, and not bigsnpr.
::::


## Example {#exo-gwas}

Let's reuse the data prepared in \@ref(exo-preprocessing) and the PCs in \@ref(exo-pca).

```{r}
library(bigsnpr)
obj.bigsnp <- snp_attach("tmp-data/GWAS_data_sorted_QC.rds")
NCORES <- nb_cores()
```


```{r}
obj.svd <- readRDS("tmp-data/PCA_GWAS_data.rds")
PC <- predict(obj.svd)
```

The clinical data includes age, sex, high-density lipoprotein (HDL)-cholesterol (`hdl`), low-density lipoprotein (LDL)-cholesterol (`ldl`), triglycerides (`tg`) and coronary artery disease status (`CAD`).

For the set of covariates, we will use sex, age, and the first 6 PCs:

```{r}
covar <- cbind(as.matrix(obj.bigsnp$fam[c("sex", "age")]), PC[, 1:6])
```

:::: {.infobox .caution}
You probably should not account for other information such as cholesterol as they are heritable covariates, which can lead to collider bias [@aschard2015adjusting; @day2016robust].
::::

```{r}
G <- obj.bigsnp$genotypes
y <- obj.bigsnp$fam$CAD
ind.gwas <- which(!is.na(y) & complete.cases(covar))
```

:::: {.infobox .caution}
To only use a subset of the data stored as an FBM (`G` here), you should almost never make a copy of the data; instead, use parameters `ind.row` (or `ind.train`) and `ind.col` to apply functions to a subset of the data.
::::

Let's perform a case-control GWAS for CAD:

```{r}
gwas <- runonce::save_run(
  big_univLogReg(G, y[ind.gwas], ind.train = ind.gwas,
                 covar.train = covar[ind.gwas, ], ncores = NCORES),
  file = "tmp-data/GWAS_CAD.rds")
```

:::: {.infobox .info}
This takes about two minutes with 4 cores on my laptop. Note that `big_univLinReg()` takes two seconds only, and should give very similar p-values, if you just need something quick.
::::

```{r, out.width="95%", fig.asp=0.5, fig.width=10, dev='png'}
plot(gwas)
CHR <- obj.bigsnp$map$chromosome
POS <- obj.bigsnp$map$physical.pos
snp_manhattan(gwas, CHR, POS, npoints = 50e3) +
  ggplot2::geom_hline(yintercept = -log10(5e-8), linetype = 2, color = "red")
```
Here, nothing is genome-wide significant because of the small sample size.

:::: {.infobox .info}
Normally, a p-value threshold of $5 \times 10^{-8}$ is used when reporting significant findings, which corresponds to correcting for one million independent tests. Now, with larger GWAS datasets that include tens of millions of variants, researchers have started using an even more stringent threshold of $5 \times 10^{-9}$.
:::: 

:::: {.infobox .exo}
Perform a GWAS for the other phenotypes, and look at the histogram, Q-Q plot and Manhattan plot.
::::

<details>
<summary>Click to see solution</summary>

```{r, out.width="95%", fig.asp=0.5, fig.width=10, dev='png'}
y2 <- obj.bigsnp$fam$hdl
ind.gwas2 <- which(!is.na(y2) & complete.cases(covar))
gwas2 <- big_univLinReg(G, y2[ind.gwas2], ind.train = ind.gwas2,
                        covar.train = covar[ind.gwas2, ], ncores = NCORES)
snp_manhattan(gwas2, CHR, POS, npoints = 50e3) +
  ggplot2::geom_hline(yintercept = -log10(5e-8), linetype = 2, color = "red")
```
</details>

***

To report independent findings (instead of all variants in one peak), people have used LD clumping, which iteratively keeps the most associated variant and remove those that are too correlated with this one. Another strategy is to use GCTA-COJO [@yang2012conditional].

:::: {.infobox .exo}
Perform some LD clumping for all variants with a p-value lower than $10^{-5}$, using a stringent $r^2$ threshold of 0.05 over a 5 Mb window. How many variants are left?
::::

<details>
<summary>Click to see solution</summary>

```{r, out.width="95%", fig.asp=0.5, fig.width=10, dev='png'}
lpval <- -predict(gwas)  # -log10(pval)
ind_keep <- snp_clumping(G, CHR, S = lpval, thr.r2 = 0.05,
                         size = 5e6, infos.pos = POS, 
                         exclude = which(lpval < -log10(1e-5)),
                         ncores = NCORES)
obj.bigsnp$map[ind_keep, ]

snp_manhattan(gwas, CHR, POS, npoints = 50e3, ind.highlight = ind_keep) +
  ggplot2::geom_hline(yintercept = -log10(5e-8), linetype = 2, color = "red") +
  ggplot2::geom_hline(yintercept = -log10(1e-5), linetype = 2, color = "blue")
```

</details>

***

For some example code to run a GWAS on multiple nodes on an HPC cluster:

- [GWAS in iPSYCH](https://github.com/privefl/bigsnpr-extdoc/blob/main/example-code/3-GWAS.R); you can perform the GWAS on multiple nodes in parallel that would each process a chunk of the variants only
- [GWAS for very large data and multiple phenotypes](https://github.com/privefl/UKBB-PGS/blob/main/code/run-GWAS-large.R); you should perform the GWAS for all phenotypes for a "small" chunk of columns to avoid repeated access from disk, and can process these chunks on multiple nodes in parallel
- [some template for {future.batchtools} when using Slurm](https://github.com/privefl/bigsnpr-extdoc/blob/main/example-code/batchtools.slurm.tmpl)


## REGENIE 

[regenie](https://rgcgithub.github.io/regenie/) [@mbatchou2021computationally] is a C++ program for whole genome regression modeling of large genome-wide association studies.
It is developed and supported by a team of scientists at the Regeneron Genetics Center.
It is fancier than simple linear/logistic regressions and is currently (one of if not) the state-of-the-art GWAS method.

REGENIE has the following properties:

- It works on both quantitative and binary traits, including binary traits with unbalanced case-control ratios

- It can handle population structure and relatedness

- It can provide some power boost
    
- It is fast and memory efficient 🔥

- It can process multiple phenotypes at once efficiently

- For binary traits, it supports Firth logistic regression and an SPA test (useful in case of very unbalanced phenotypes)

- It can perform gene/region-based tests (burden, SBAT, SKAT/SKATO, ACATV/ACATO)

- It can perform interaction tests (GxE, GxG) as well as conditional analyses

- Meta-analysis of REGENIE summary statistics can be performed using REMETA
    
- It supports the BGEN, PLINK bed/bim/fam and PLINK2 pgen/pvar/psam genetic data formats

- It can be installed with Conda


## Genomic control and LD score regression

$\lambda_\text{GC}$, the genomic inflation factor (GIF), was used a lot to assess whether a GWAS is confounded by population structure or relatedness (when e.g. larger than 1.05).
It corresponds to the ratio between the median of test statistics over the median of expected test statistics (under the null hypothesis).
However, it is no longer used because, with sufficient GWAS power and trait polygenicity, a GIF larger than 1 is actually expected [@yang2011genomic].

```{r}
snp_qq(gwas) + 
  ggplot2::ylim(1, NA)  # to plot less points
```

***

Then, the intercept of LD score regression was introduced to test for confounding [@bulik2015ld]. 
However, note that it was observed that LD score regression intercepts tend to increase with SNP-heritability and GWAS sample size [@loh2018mixed].
