# Genome-Wide Association Study (GWAS)

```{r setup, include=FALSE}
WORDS_TO_IGNORE <- c("gwas", "HDL", "https", "LDL", "lipoprotein", "batchtools", "Slurm", "ACATO", "ACATV", "Conda", "GxE", "GxG", "Regeneron", "regenie", "REGENIE", "REMETA", "SBAT", "SKAT", "SKATO")
source("knitr-options.R")
source("spelling-check.R")
```

In {bigstatsr}, you can perform both standard linear and logistic regressions GWAS, using either `big_univLinReg()` or `big_univLogReg()`.
Function `big_univLinReg()` should be very fast, while `big_univLogReg()` is slower.

:::: {.infobox .info}
This type of association, where each variable is considered independently, can be performed for any type of FBM (i.e. it does not have to be a genotype matrix). This is why these two functions are in package {bigstatsr}, and not {bigsnpr}.
::::


## Example {#exo-gwas}

Let us reuse the data prepared in \@ref(exo-preprocessing) and the PCs in \@ref(exo-pca).

```{r}
library(bigsnpr)
obj.bigsnp <- snp_attach("tmp-data/GWAS_data_sorted_QC.rds")
NCORES <- nb_cores()
```


```{r}
obj.svd <- readRDS("tmp-data/PCA_GWAS_data.rds")
PC <- predict(obj.svd)
```

The clinical data includes age, sex, high-density lipoprotein (HDL)-cholesterol (`hdl`), low-density lipoprotein (LDL)-cholesterol (`ldl`), triglycerides (`tg`) and coronary artery disease status (`CAD`).

For the set of covariates, we will use sex, age, and the first 6 PCs:

```{r}
covar <- cbind(as.matrix(obj.bigsnp$fam[c("sex", "age")]), PC[, 1:6])
```

:::: {.infobox .caution}
You probably should not account for other information such as cholesterol as they are heritable covariates, which can lead to collider bias [@aschard2015adjusting; @day2016robust].
::::

```{r}
G <- obj.bigsnp$genotypes
y <- obj.bigsnp$fam$CAD
ind.gwas <- which(!is.na(y) & complete.cases(covar))
```

:::: {.infobox .caution}
To only use a subset of the data stored as an FBM (`G` here), you should almost never make a copy of the data; instead, use parameters `ind.row` (or `ind.train`) and `ind.col` to apply functions to a subset of the data.
::::

Let us perform a case-control GWAS for CAD:

```{r}
gwas <- runonce::save_run(
  big_univLogReg(G, y[ind.gwas], ind.train = ind.gwas,
                 covar.train = covar[ind.gwas, ], ncores = NCORES),
  file = "tmp-data/GWAS_CAD.rds")
```

:::: {.infobox .info}
This takes about two minutes with 4 cores on my laptop. Note that `big_univLinReg()` takes two seconds only, and should give very similar p-values, if you just need something quick.
::::

```{r, out.width="95%", fig.asp=0.5, fig.width=10, dev='png'}
plot(gwas)
CHR <- obj.bigsnp$map$chromosome
POS <- obj.bigsnp$map$physical.pos
snp_manhattan(gwas, CHR, POS, npoints = 50e3) +
  ggplot2::geom_hline(yintercept = -log10(5e-8), linetype = 2, color = "red")
```
Here, nothing is genome-wide significant because of the small sample size.


:::: {.infobox .exo}
Perform a GWAS for the other phenotypes, and look at the histogram, Q-Q plot and Manhattan plot.
::::

<details>
<summary>Click to see solution</summary>

```{r, out.width="95%", fig.asp=0.5, fig.width=10, dev='png'}
y2 <- obj.bigsnp$fam$hdl
ind.gwas2 <- which(!is.na(y2) & complete.cases(covar))
gwas2 <- big_univLinReg(G, y2[ind.gwas2], ind.train = ind.gwas2,
                        covar.train = covar[ind.gwas2, ], ncores = NCORES)
snp_manhattan(gwas2, CHR, POS, npoints = 50e3) +
  ggplot2::geom_hline(yintercept = -log10(5e-8), linetype = 2, color = "red")
```

<details>

***

Some other example code:

- [GWAS in iPSYCH](https://github.com/privefl/bigsnpr-extdoc/blob/main/example-code/3-GWAS.R); you can perform the GWAS on multiple nodes in parallel that would each process a chunk of the variants only
- [GWAS for very large data and multiple phenotypes](https://github.com/privefl/UKBB-PGS/blob/main/code/run-GWAS-large.R); you should perform the GWAS for all phenotypes for a "small" chunk of columns to avoid repeated access from disk, and can process these chunks on multiple nodes in parallel
- [some template for {future.batchtools} when using Slurm](https://github.com/privefl/bigsnpr-extdoc/blob/main/example-code/batchtools.slurm.tmpl)


## REGENIE 

[regenie](https://rgcgithub.github.io/regenie/) [@mbatchou2021computationally] is a C++ program for whole genome regression modeling of large genome-wide association studies.

It is developed and supported by a team of scientists at the Regeneron Genetics Center.

The method has the following properties:

- It works on quantitative and binary traits, including binary traits with unbalanced case-control ratios

- It can handle population structure and relatedness

- It can process multiple phenotypes at once efficiently

- For binary traits, it supports Firth logistic regression and an SPA test

- It can perform gene/region-based tests (Burden, SBAT, SKAT/SKATO, ACATV/ACATO)

- It can perform interaction tests (GxE, GxG) as well as conditional analyses

- Meta-analysis of REGENIE summary statistics can be performed using REMETA
    
- It is fast and memory efficient 🔥

- It can provide some power boost
    
- It supports the BGEN, PLINK bed/bim/fam and PLINK2 pgen/pvar/psam genetic data formats

- It can be installed with Conda
